The Companion Mass Distribution to the ELM WD Sample

### 2014-10-02 documents/letter/ms.tex
acknowledgements and shiii

We construct a statistical model to derive constraints on a parametric model --- parametrized by the vector $\bs{\theta}$\footnote{In what follows, vectors or sets of parameters or quantites are represented by bold symbols.} --- for the distribution of ELM WD companion masses, $p(M_2 \given \bs{\theta})$. For each system, we assume we are given $K$, $T$, and $M_1$ and therefore know the value of the mass function, $m_f$ (Eq.~\ref{eq:massfunc}). We would like to find posterior constraints on the parameters, $\boldsymbol{\theta}$, given all of the observed mass functions, $\bs{m}_f$, by deconvolving the distribution of mass functions from the unobserved inclination angle, $i$. Using Bayes' rule,
\begin{equation}
    p(\bs{\theta} \given m_f) = \frac{1}{\mathcal{Z}}~p(m_f \given \bs{\theta})~p(\bs{\theta}).
\end{equation}
where $p(m_f \given \bs{\theta})$ is the likelihood, $p(\bs{\theta})$ is the prior on parameters $\bs{\theta}$, and the evidence integral, $\mathcal{Z}$, is a constant that depends only on the data. The marginal likelihood, $p(m_f \given \bs{\theta})$, involves integrals over the latent variables $i$ and companion mass, $M_2$,
\begin{equation}
    p(m_f \given \bs{\theta}) = \int_0^{\pi/2} di \int_{-\infty}^\infty dM_2~p(m_f \given M_1, M_2, i)~p(M_2 \given \bs{\theta})~p(i).
\end{equation}
where we neglect observational uncertainties in the mass function,\footnote{This is justified because the relative uncertainties in these quantities are small, $\sigma_x / x \sim 0.1$, however in future work we will extend this methodology to include uncertainties in all measured quantities.} and assume the inclination angles, $i$, are isotropically distributed
\begin{align}
	p(m_f \given M_1, M_2, i) &= \delta(m_f - f(M_1, M_2, i))\\
	f(M_1, M_2, i) &= \frac{(M_2 \sin i)^3}{(M_1 + M_2)^2}\\
	p(i) &= \sin i.
\end{align}
For now, we keep this general by not yet specifying a parametric form for the companion mass distribution, $p(M_2 \given \bs{\theta})$. With the above assumptions, the marginal likelihood integral is
\begin{align}
    p(m_f \given \bs{\theta}) &= \int_{0}^\infty dM_2~p(M_2 \given \bs{\theta}) \int_0^{\pi/2} di ~ \delta(m_f - f(M_1, M_2, i))~\sin i\\
    &= \int_{0}^\infty dM_2~p(M_2 \given \bs{\theta}) \int_0^{\pi/2} di ~\sin i ~ \delta(g(M_1,M_2,i))\label{eq:delta}
\end{align}
where
\begin{equation}
	g(M_1,M_2,i) = \frac{M_2^3}{(M_1+M_2)^2}\sin^3 i - m_f.
\end{equation}
The inner integral (over $i$) has the form
\begin{equation}
    \int dx~F(x)~\delta{(G(x)}) = \sum_j \frac{F(x^*_j)}{|G'(x^*_j)|}
\end{equation}
where the sum is over the roots, $x^*_j$, of the function $G(x)$. The root, $i^*$, and derivative of the argument of the delta function in Equation~\ref{eq:delta} are 
\begin{align}
	i^* &= \sin^{-1}\alpha\\
	\alpha &= \frac{(m_f(M_1+M_2)^2)^{1/3}}{M_2}\\
	\frac{\partial g}{\partial i} &= \frac{3M_2^3}{(M_1+M_2)^2}\sin^2 i \cos i\\
	\frac{\partial g}{\partial i}\bigg\rvert_{i^*} &= \frac{3M_2^3}{(M_1+M_2)^2} \alpha^2 \sqrt{1 - \alpha^2}
\end{align}
With this in mind, we may write the marginal likelihood as
\begin{align}
	p(m_f \given \bs{\theta}) &= \int_{0}^\infty dM_2~p(M_2 \given \bs{\theta})~\frac{\sin i^*}{g'(M_1,M_2,i^*)}\\
	&= \int_{0}^\infty dM_2~p(M_2 \given \bs{\theta})~\alpha\left[\frac{3M_2^3}{(M_1+M_2)^2} \alpha^2 \sqrt{1 - \alpha^2}\right]^{-1}\\
	&= \int_{M_{2,{\rm min}}}^\infty dM_2~p(M_2 \given \bs{\theta})~\frac{(M_1+M_2)^{4/3}}{3m_f^{1/3}M_2\sqrt{M_2^2 - (m_f(M_1+M_2)^2)^{2/3}}}\label{eq:fullm2}
\end{align}
where the bottom bound in the integral in Equation~\ref{eq:fullm2} is the minimum companion mass for which the integrand is real, $M_{2,{\rm min}}$.

\vspace{2em}
\hrule

To determine the individual system likelihoods, we first marginalize over $M_2$:

\begin{equation}
P(D|\mathbf{\Theta}) = \frac{1}{A} \int_0^{\infty} P(D |\mathbf{\Theta}, M_2) P	(M_2|\mathbf{\Theta})\ dM_2 
\end{equation}

where $A$ is a normalization constant. We further marginalize over the inclination angle, $i$:

\begin{equation}
P(D|\mathbf{\Theta}) = \frac{1}{A} \int_0^{\pi/2} \int_{M_{2,min}}^{\infty} P(D | \mathbf{\Theta}, M_2, i) P(M_2|\mathbf{\Theta}, i) P(i | \mathbf{\Theta})\ {\rm d}M_2\ {\rm d}i
\end{equation}

Since only a single value for $M_{2,min}$ is allowed for a given $M_1$, $M_2$, and $i$, $P(D | \mathbf{\Theta}, M_2, i)$ reduces to a delta function:

\begin{equation}
P(D | \mathbf{\Theta}, M_2, i) = \delta \left(M_{2,min} - f(M_1,M_2,i) \right)
\end{equation}

where $f(M_1,M_2,i)$ is the root of equation {\bf XXXXX}. Furthermore, the inclination angle is independent of other parameters:

\begin{eqnarray}
P(i | \mathbf{\Theta}) &=& P(i) \\
  &=& {\rm sin}\ i \\
  &=& \left( \frac{M_1 + M_2}{M_1 + M_{2,min}} \right)^{2/3} \frac{M_{2,min}}{M_2}
\end{eqnarray}

where in the last equation, we used equation {\bf XXX}. Combining equations {\bf XXX}, {\bf XXX}, and {\bf XXX} and evaluating the integral over $i$ we produce the following result:

\begin{equation}
P(D|\mathbf{\Theta}) = \frac{1}{A} \int_{M_{2,min}}^{\infty}  P(M_2|\mathbf{\Theta},i_{M_2}) \left( \frac{M_1 + M_2}{M_1 + M_{2,min}} \right)^{2/3} \frac{M_{2,min}}{M_2} {\rm d}M_2
\end{equation}

where $i_{M_2}$ is a function of $M_2$, determined from Equation {\bf XXX}. The final parameter left to be determined is the normalization constant, $A$. This parameter requires an additional integral, over all $M_{2,min}$:

\begin{equation}
A = \int_0^{\infty} \int_{M_{2,min}}^{\infty} P(M_2|\mathbf{\Theta}, i_{M_2}) \left( \frac{M_1 + M_2}{M_1 + M_{2,min}} \right)^{2/3} \frac{M_{2,min}}{M_2} {\rm d}M_2\ {\rm d}M_{2,min}
\end{equation}

Finally, $P(M_2|\mathbf{\Theta},i_{M_2})$ is our model. We choose to model the masses of the WD companions as a Gaussian, with an additional NS component. This model is independent of the inclination angle and can be expressed in the following way:

\begin{eqnarray}
P(M_2|\mathbf{\Theta},i) &=& P(M_2 | \mathbf{\Theta}) \\
  &=& \left( 1-\mathcal{f}_{NS} \right) \mathcal{N}(\mu , \sigma)|_{M_2} + \mathcal{f}_{NS} \delta \left( M_2 - M_{NS} \right)
\end{eqnarray}

where $\mathcal{f}_{NS}$ represents the NS fraction in the model, and $\mathcal{N}(\mu , \sigma)$ represents a Gaussian with mean, $\mu$, and standard deviation, $\sigma$. The NS component is modeled by a delta function around $M_{NS}$ which we take to be $1.35 ~\Msun$. Combining this with equation {\bf XXX}, we can separate the integrand into a WD and a NS component:

\begin{equation}
P(D|\mathbf{\Theta}) = P_{WD} + P_{NS}
\end{equation}

where the WD component is:

\begin{equation}
P_{WD} = \frac{1}{A} \int_{M_{2,min}}^{\infty} \left[ \left( 1-\mathcal{f}_{NS} \right) \mathcal{N}(\mu , \sigma)|_{M_2} \right]  \left( \frac{M_1 + M_2}{M_1 + M_{2,min}} \right)^{2/3} \frac{M_{2,min}}{M_2} {\rm d}M_2 
\end{equation}

and the NS component is:

\begin{equation}
P_{NS} = \frac{1}{A} \mathcal{f}_{NS}\left( \frac{M_1 + M_{NS}}{M_1 + M_{2,min}} \right)^{2/3} \frac{M_{2,min}}{M_{NS}}
\end{equation}

The normalization constant can likewise be reduced into a WD and a NS component, with the integral over d$M_{2,min}$ performed separately on both:

\begin{equation}
A = A_{WD} + A_{NS}
\end{equation}

where the WD component is:

\begin{equation}
A_{WD} = \int_0^{\infty} \int_{M_{2,min}}^{\infty} \left[ \left( 1-\mathcal{f}_{NS} \right) \mathcal{N}(\mu , \sigma)|_{M_2} \right]  \left( \frac{M_1 + M_2}{M_1 + M_{2,min}} \right)^{2/3} \frac{M_{2,min}}{M_2}\ {\rm d}M_2\ {\rm d}M_{2,min}
\end{equation}

and the NS component is:

\begin{equation}
A_{NS} = \int_0^{\infty} \mathcal{f}_{NS}\left( \frac{M_1 + M_{NS}}{M_1 + M_{2,min}} \right)^{2/3} \frac{M_{2,min}}{M_2} \delta \left( M_2 - M_{NS} \right) {\rm d}M_{2,min}
\end{equation}

which can be evaluated analytically, using integration by parts:

\begin{equation}
A_{NS} = 3 \left(M_1+M_{NS}\right)^{2/3} \left[  \frac{M_{2,min}}{M_{NS}} \left( M_1+M_{2,min} \right)^{1/3} - \frac{3}{4} \frac{ \left( M_1 + M_{2,min} \right)^{4/3} }{M_{NS}} \right] |_{M_{2,min}=0}^{M_{2,min}=M_{NS}}
\end{equation}

Using these equations, we can now calculate the likelihood of a given set of model parameters ($\mu$, $\sigma$, and $\mathcal{f}_{NS}$). In general, these integrals must be evaluated numerically. Furthermore, since each system has a different $M_1$ and $M_{2,min}$ and therefore must be calculated separately, these formulae become quite computationally expensive, particularly calculating the double integral in $A_{WD}$.

It is also straightforward to determine, for a particular $\Theta$, the probability of any individual system being composed of a NS:

\begin{equation}
P(NS) = \frac{P_{NS}}{P_{WD} + P_{NS}}
\end{equation}

where $P_{NS}$ and $P_{WD}$ are calculated from equation {\bf XXX} and {\bf XXX}, respectively.

@@

\acknowledgements
APW is supported by a National Science Foundation Graduate Research Fellowship under Grant No.\ 11-44155. 
This research made use of Astropy, a community-developed core \texttt{Python} package for Astronomy \citep{astropy13}.
This work additionally relied on Columbia University's \emph{Hotfoot} and \emph{Yeti} compute clusters, and we acknowledge the Columbia HPC support staff for assistance. \\

>>>>

We construct a statistical model to derive constraints on a parametric model, parametrized by the vector $\bs{\theta}$,\footnote{In what follows, vectors or sets of parameters or quantites are represented by bold symbols.} for the distribution of ELM WD companion masses, $p(M_2 \given \bs{\theta})$. For each system, we assume we are given $K$, $T$, and $M_1$ and therefore know the value of the mass function, $m_f$ (Eq.~\ref{eq:massfunc}). We would like to find posterior constraints on the parameters, $\boldsymbol{\theta}$, given all of the observed mass functions, $\bs{m}_f$, by deconvolving the distribution of mass functions from the unobserved inclination angle, $i$. Using Bayes' rule,
\begin{equation}
    p(\bs{\theta} \given m_f) = \frac{1}{\mathcal{Z}}~p(m_f \given \bs{\theta})~p(\bs{\theta}).
\end{equation}
where $p(m_f \given \bs{\theta})$ is the likelihood, $p(\bs{\theta})$ is the prior on parameters $\bs{\theta}$, and the evidence integral, $\mathcal{Z}$, is a constant that depends only on the data. The marginal likelihood, $p(m_f \given \bs{\theta})$, involves integrals over the latent variables $i$ and companion mass, $M_2$,
\begin{equation}
    p(m_f \given \bs{\theta}) = \int_0^{\pi/2} di \int_{-\infty}^\infty dM_2~p(m_f \given M_1, M_2, i)~p(M_2 \given \bs{\theta})~p(i).
\end{equation}
where we neglect observational uncertainties in the mass function,\footnote{This is justified because the relative uncertainties in these quantities are small, $\sigma_x / x \sim 0.1$, however in future work we will extend this methodology to include uncertainties in all measured quantities.} and assume the inclination angles, $i$, are isotropically distributed
\begin{align}
	p(m_f \given M_1, M_2, i) &= \delta(m_f - f(M_1, M_2, i))\\
	f(M_1, M_2, i) &= \frac{(M_2 \sin i)^3}{(M_1 + M_2)^2}\\
	p(i) &= \sin i.
\end{align}
For now, we keep this general by not yet specifying a parametric form for the companion mass distribution, $p(M_2 \given \bs{\theta})$. With the above assumptions, the marginal likelihood integral is
\begin{align}
    p(m_f \given \bs{\theta}) &= \int_{0}^\infty dM_2~p(M_2 \given \bs{\theta}) \int_0^{\pi/2} di ~ \delta(m_f - f(M_1, M_2, i))~\sin i\\
    &= \int_{0}^\infty dM_2~p(M_2 \given \bs{\theta}) \int_0^{\pi/2} di ~\sin i ~ \delta(g(M_1,M_2,i))\label{eq:delta}
\end{align}
where
\begin{equation}
	g(M_1,M_2,i) = \frac{M_2^3}{(M_1+M_2)^2}\sin^3 i - m_f.
\end{equation}
The inner integral (over $i$) has the form
\begin{equation}
    \int dx~F(x)~\delta{(G(x)}) = \sum_j \frac{F(x^*_j)}{|G'(x^*_j)|}
\end{equation}
where the sum is over the roots, $x^*_j$, of the function $G(x)$. The root, $i^*$, and derivative of the argument of the delta function in Equation~\ref{eq:delta} are 
\begin{align}
	i^* &= \sin^{-1}\alpha\\
	\alpha &= \frac{(m_f(M_1+M_2)^2)^{1/3}}{M_2}\\
	\frac{\partial g}{\partial i} &= \frac{3M_2^3}{(M_1+M_2)^2}\sin^2 i \cos i\\
	\frac{\partial g}{\partial i}\bigg\rvert_{i^*} &= \frac{3M_2^3}{(M_1+M_2)^2} \alpha^2 \sqrt{1 - \alpha^2}
\end{align}
With this in mind, we may write the marginal likelihood as
\begin{align}
	p(m_f \given \bs{\theta}) &= \int_{0}^\infty dM_2~p(M_2 \given \bs{\theta})~\frac{\sin i^*}{g'(M_1,M_2,i^*)}\\
	&= \int_{0}^\infty dM_2~p(M_2 \given \bs{\theta})~\alpha\left[\frac{3M_2^3}{(M_1+M_2)^2} \alpha^2 \sqrt{1 - \alpha^2}\right]^{-1}\\
	&= \int_{M_{2,{\rm min}}}^\infty dM_2~p(M_2 \given \bs{\theta})~\frac{(M_1+M_2)^{4/3}}{3m_f^{1/3}M_2\sqrt{M_2^2 - (m_f(M_1+M_2)^2)^{2/3}}}\label{eq:fullm2}
\end{align}
where the bottom bound in the integral in Equation~\ref{eq:fullm2} is the minimum companion mass for which the integrand is real, $M_{2,{\rm min}}$.

[TODO: put in $p(M_2\given ...), description of models in separate subsection?]

@@

\acknowledgements
The authors wish to acknowledge David Hogg (NYU) for useful discussions, and the organizing committee of the \emph{AstroData Hack Week} (2014). 
APW is supported by a National Science Foundation Graduate Research Fellowship under Grant No.\ 11-44155. 
This research made use of Astropy, a community-developed core \texttt{Python} package for Astronomy \citep{astropy13}.
This work additionally relied on Columbia University's \emph{Hotfoot} and \emph{Yeti} compute clusters, and we acknowledge the Columbia HPC support staff for assistance. \\
